---
title: "00 - Data Setup"
format: html
editor: source
execute:
  echo: true
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

## Under 75 respiratory mortality surveillance data

```{r}
#| label: setup
#| include: false

library(fingertipsR)
library(tidyverse)

# Find the indicator ID for under 75 respiratory mortality
all_indicators <- indicators()

# Filter to find respiratory mortality indicators
respiratory_inds <- all_indicators |>  
  filter(grepl("Under 75 mortality.*respiratory", IndicatorName, ignore.case = TRUE))

# View to choose the right one
# View(respiratory_inds)

indicator_id <- 40701

# Download fresh data
raw_data <- fingertips_data(IndicatorID = indicator_id, AreaTypeID = 6)

# Clean and save
respiratory_data <- raw_data |> 
  select(
    region = AreaName,
    time_period = Timeperiod,
    value = Value,
    count = Count,
    denominator = Denominator,
    lower_ci = LowerCI95.0limit,
    upper_ci = UpperCI95.0limit,
    sex = Sex,
    age = Age
  ) |> 
  mutate(year = as.numeric(str_extract(time_period, "\\d{4}")))

# Save
saveRDS(respiratory_data, "data/respiratory_data.rds")
print("Data ready.")
```

The `time_period` column contains both single-year estimates ("2024") and three‑year averages ("2022 - 24"). To enable chronological plotting, I extracted the first four digits to create a numeric `year` column. For three‑year ranges, I assigned the value to the start year: a practical choice that maintains time order and allows these period averages to be visualised alongside single‑year estimates without overlap. The original time period is preserved in the data should more detail be needed.

## C. Difficile Data for Outbreak Detection

```{r}
#| label: download-cdiff-data
#| include: false

library(ukhsadatR)

# Create a vector of all UKHSA regions
ukhsa_regions <- c(
  "East Midlands",
  "East of England",
  "London",
  "North East",
  "North West",
  "South East",
  "South West",
  "West Midlands",
  "Yorkshire and Humber"
)

# Download C. difficile data for each region and combine
cdiff_list <- list()

for(region in ukhsa_regions) {
  message("Downloading data for: ", region)
  region_data <- get_data(
    theme = "infectious_disease",
    sub_theme = "gastrointestinal",
    topic = "C-difficile",
    geography_type = "UKHSA Region",
    geography = region,
    metric = "c-difficile_cases_countsByOnsetType"
)

  if(!is.null(region_data)) {
    cdiff_list[[region]] <- region_data
  }
}

# Combine all regions
cdiff_raw <- bind_rows(cdiff_list)

# Clean and C. Difficile save
cdiff_data <- cdiff_raw |> 
  mutate(
    date = as.Date(date),
    week = epiweek,  # Use the epiweek column from the data
    year = year,  # Already present
    region = as.factor(geography),  # geography column contains region names
    onset_type = stratum,  # Rename for clarity (Community-onset, Hospital-onset, etc.)
    cases = metric_value,  # Rename to match our analysis code
    id = paste(region, year, sprintf("%02d", week), sep = "_")
  ) |> 
  select(
    region, year, week, date, onset_type, cases,
    geography_code, metric, sex, age, everything()
  ) |> 
  filter(!is.na(cases)) |> 
  arrange(region, year, week, onset_type)

# Check the cleaned data
glimpse(cdiff_data)

# Save to RDS
saveRDS(cdiff_data, "data/cdiff_data.rds")

print("C. difficile data downloaded and saved successfully.")
print(paste("Total rows:", nrow(cdiff_data)))
print(paste("Regions:", paste(unique(cdiff_data$region), collapse = ", ")))
print(paste("Years:", paste(unique(cdiff_data$year), collapse = ", ")))
print(paste("Onset types:", paste(unique(cdiff_data$onset_type), collapse = ", ")))
```

## MMR Vaccination Coverage Data

```{r}
#| label: mmr-indicator
#| include: false

# Find MMR indicator ID
all_indicator <- indicators()
mmr_indicator <- all_indicator |> 
  filter(grepl("MMR", IndicatorName, ignore.case = TRUE))
View(all_indicator)
```

```{r}
#| label: preliminary-exploration
#| include: false

library(fingertipsR)
library(tidyverse)

# Get MMR data for all unitary authorities/counties
mmr_explore <- fingertips_data(
  IndicatorID = 30309,
  AreaTypeID = 502
)

# See what areas are available
area_summary <- mmr_explore |> 
  group_by(AreaName) |> 
  summarise(
    years = n_distinct(Timeperiod),
    latest_value = Value[which.max(as.numeric(str_extract(Timeperiod, "\\d{4}")))],
    .groups = "drop"
  ) |> 
  arrange(desc(years))

View(area_summary)

# Check specific cities you're interested in
cities_of_interest <- c("Nottingham", "Leicester", "Bristol", "York", "Derby", "Peterborough", "Birmingham")

mmr_explore |>
  filter(AreaName %in% cities_of_interest) |>
  select(AreaName, Timeperiod, Value) |>
  arrange(AreaName, Timeperiod)

# Filter to your shortlist
cities_to_plot <- c("Leicester", "Nottingham", "Bristol", "Birmingham", "York", "Derby", "Peterborough")

mmr_cities <- mmr_explore |>
  filter(AreaName %in% cities_to_plot)

# Create the plot
ggplot(mmr_cities, aes(x = Timeperiod, y = Value, color = AreaName, group = AreaName)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_color_viridis_d() +
  labs(
    title = "MMR coverage trends in selected cities",
    subtitle = "2010/11 to 2024/25",
    x = "Year",
    y = "Coverage (%)",
    color = "City",
    caption = "Data: Fingertips (Indicator 30309)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
#| label: download-mmr
#| include: false

library(fingertipsR)
library(tidyverse)

# Download MMR data using correct AreaTypeIDs
# Indicator 30309: MMR for one dose at age 2

# Leicester City (Unitary Authority - AreaTypeID: 502)
mmr_leicester <- fingertips_data(
  IndicatorID = 30309,
  AreaTypeID = 502
) |> 
  filter(AreaName == "Leicester")

# England (AreaTypeID: 15)
mmr_england <- fingertips_data(
  IndicatorID = 30309,
  AreaTypeID = 15
) |> 
  filter(
    AreaName == "England")

# East Midlands authorities for regional average
em_raw <- fingertips_data(
  IndicatorID = 30309,
  AreaTypeID = 502
) |>
  filter(
    AreaName %in% c(
      "Derby", "Leicester", "Lincolnshire", "Northamptonshire", 
      "Nottingham", "Nottinghamshire", "Rutland", "Leicestershire"
    )
  )

# Calculate East Midlands average
mmr_em_average <- em_raw |>
  group_by(Timeperiod) |>
  summarise(
    coverage = mean(Value, na.rm = TRUE),
    geography = "East Midlands (average)",
    .groups = "drop"
  ) |>
  mutate(
    year = as.numeric(str_extract(Timeperiod, "\\d{4}")),
    coverage = round(coverage, 1)
  )

# Combine main datasets
mmr_data <- bind_rows(
  mmr_leicester |> select(geography = AreaName, time_period = Timeperiod, coverage = Value),
  mmr_england |> select(geography = AreaName, time_period = Timeperiod, coverage = Value),
  mmr_em_average |> select(geography, time_period = Timeperiod, coverage)
) |>
  mutate(
    year = as.numeric(str_extract(time_period, "\\d{4}")),
    coverage = round(coverage, 1)
  ) |>
  filter(year >= 2010) |>
  arrange(geography, year)

# Save main data
saveRDS(mmr_data, "data/mmr_data.rds")

# Save benchmark data separately
saveRDS(em_raw, "data/em_authorities.rds")

print("MMR coverage data downloaded and saved successfully!")
print(paste("Leicester years:", nrow(mmr_leicester)))
print(paste("East Midlands average years:", nrow(mmr_em_average)))
print(paste("England years:", nrow(mmr_england)))
```

## North-South Grouped Respiratory Data (for Day 4)

```{r}
#| label: prepare-north-south
#| include: false

library(tidyverse)
library(stringr)

# Load the saved respiratory data from day 1
respiratory_data <- readRDS("data/respiratory_data.rds")

# Define region groupings
north_regions <- c(
    "North East region (statistical)",
  "North West region (statistical)",
  "Yorkshire and the Humber region (statistical)"
  )
south_regions <- c(
    "South East region (statistical)",
  "South West region (statistical)",
  "London region (statistical)",
  "East of England region (statistical)"
  )
excluded_regions <- c(
    "East Midlands region (statistical)",
  "West Midlands region (statistical)",
  "England"  # Exclude national level
  )

# Add grouping variable and filter to North/South only
north_south_data <- respiratory_data |> 
  mutate(
    region_group = case_when(
      region %in% north_regions ~ "North",
      region %in% south_regions ~ "South",
      TRUE ~ "Excluded"
    )
  ) |> 
  filter(region_group != "Excluded") |> 
  select(region, region_group, year, mortality_rate = value, deaths = count)

glimpse(north_south_data)

# Verify regions are correctly assigned
north_south_data |> 
  distinct(region, region_group) |> 
  arrange(region_group, region) |> 
  print()

# Save for Day 4
saveRDS(north_south_data, "data/north_south_data.rds")

print("North‑South grouped data saved successfully!")
print(paste("North observations:", nrow(filter(north_south_data, region_group == "North"))))
print(paste("South observations:", nrow(filter(north_south_data, region_group == "South"))))
```
