---
title: "Respiratory Disease and Immunisation in England: Key Findings"
author: "Marco Sorbona, PhD"
date: "March 2026"
format: 
  html:
    toc: false
    code-fold: false
    theme: cosmo
    css: briefing.css
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggplot2)
library(scales)
library(knitr)

# Load data
respiratory_data <- readRDS("data/respiratory_data.rds") |> filter(sex == "Persons")
north_south_data <- readRDS("data/north_south_data.rds")
mmr_data <- readRDS("data/mmr_data.rds")
```

## Background

This briefing summarises analysis on respiratory mortality and MMR vaccination coverage in England. Data are drawn from the **Fingertips public health database** (Indicators 40701 and 30309) and the **UKHSA dashboard**, covering 2001 - 2024.

The analysis was structured as a simulated week of work in UKHSA Field Services, responding to a local authority request and identifying regional inequalities and outbreak signals.

------------------------------------------------------------------------

## Key Findings

| Finding | Detail |
|--------------------------------------|----------------------------------|
| **North‑South mortality gap** | Northern regions have 14.1 more respiratory deaths per 100,000 than the South: a **48% higher rate** \[1\]. The gap has persisted for over two decades. |
| **Statistical significance** | p \< 0.001; Cohen's d = 2.4 (large effect). |
| **Leicester MMR decline** | Coverage fell from 95.8% (2013) to 88.4% (2024), an **8‑point drop**. Leicester has been below the WHO 95% herd immunity threshold for five consecutive years \[2\]. |
| **C. difficile outbreak signal** | East Midlands showed **80% of weeks exceeding** the Poisson threshold, warranting investigation. |

------------------------------------------------------------------------

## Interpretation

### North - South Respiratory Mortality Gap

The North has 14.1 additional respiratory deaths per 100,000 compared to the South: a 48% higher rate \[1\]. Asthma + Lung UK's *ICS Respiratory Review* shows that rates of emergency admissions and deaths from lung conditions are much higher in northern areas such as Blackpool and Liverpool, where death rates are over **two times higher** than in more affluent areas like Richmond in London \[3\].

Six out of the 10 areas with the highest rates of emergency admissions and deaths from lung conditions are in the Northwest of England \[1\]. The drivers are structural: higher smoking rates, worse air pollution, poorer housing, and less access to early diagnosis \[6\].

The gap has persisted for over two decades. A Wellcome‑funded project notes that *"there are deep-rooted and long-term regional health inequalities across England, with a two-year gap between the north and the south in how long people are expected to live"* \[5\].

```{r}
#| echo: false
#| fig-cap: "North‑South respiratory mortality gap (2001–2024)"

# Calculate annual means by group
annual_group <- north_south_data |>
  group_by(region_group, year) |>
  summarise(mean_rate = mean(mortality_rate, na.rm = TRUE), .groups = "drop")

ggplot(annual_group, aes(x = year, y = mean_rate, color = region_group)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("North" = "#e74c3c", "South" = "#3498db")) +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  labs(
    x = "Year",
    y = "Rate per 100,000",
    color = ""
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Leicester MMR Decline

Leicester's MMR coverage declined from 95.8% in 2013 to 88.4% in 2024. This pattern is consistent with regional trends: from 2013/14 to 2023/24, MMR1 coverage in five-year-olds decreased by 7.1 percentage points to 87.0% in Birmingham and by 3.5 percentage points to 91.9% in the West Midlands \[2\].

The city has been below the WHO 95% herd immunity threshold for five consecutive years. According to UKHSA data, *"no local authority in the West Midlands had coverage exceeding the WHO target of 95% in all the routine childhood immunisations"* \[2\].

This decline is significant because measles outbreaks can occur when coverage falls below 90–95% \[4\]. The WHO recommends 95% coverage for both MMR doses to achieve and maintain herd immunity. In the UK, measles elimination status was lost in 2019 following a resurgence of cases, highlighting the importance of maintaining high vaccination coverage across all communities.

```{r}
#| echo: false
#| fig-cap: "MMR coverage: Leicester vs East Midlands vs England"

mmr_key <- mmr_data |>
  filter(geography %in% c("Leicester", "East Midlands (average)", "England"))

ggplot(mmr_key, aes(x = year, y = coverage, color = geography)) +
  geom_hline(yintercept = 95, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c(
    "Leicester" = "#3498db",
    "East Midlands (average)" = "#e67e22",
    "England" = "#2c3e50"
  )) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(80, 100)) +
  labs(
    x = "Year",
    y = "Coverage (%)",
    color = ""
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### C. difficile Outbreak Signal

Using a Poisson threshold for sparse data (mean + 1.96 × √mean), the East Midlands showed **80% of weeks** exceeding expected case counts. This level of sustained exceedance warrants line list requests from local trusts and further investigation.

------------------------------------------------------------------------

## Recommendations

| Priority | Action |
|----------------------------------------|--------------------------------|
| **1** | **Target resources to northern regions**: additional investment in respiratory health services, smoking cessation, and housing improvement \[1\]. Asthma + Lung UK calls for a **£40 million spirometry recovery fund** with more funding to ICSs experiencing higher levels of deprivation \[6\]. |
| **2** | **Investigate Leicester's MMR decline**: understand local factors behind the 8‑point drop, including vaccine confidence, access, and second‑dose uptake. The government's 10‑year health plan should address childhood immunisation gaps \[2\]. |
| **3** | **Follow up C. difficile outbreak signal**: request line lists from East Midlands trusts and conduct further analysis. |
| **4** | **Monitor inequalities quarterly**: track the North‑South gap in routine surveillance \[5\]. |

------------------------------------------------------------------------

## Limitations

-   Respiratory mortality data exclude COVID‑19 deaths (coded separately)
-   MMR data are for age 2 only; coverage at age 5 may differ
-   C. difficile baseline limited to 5 weeks in 2023
-   Regional averages mask within‑area variation

------------------------------------------------------------------------

## 

## References

1.  Asthma + Lung UK. (2025). *ICS Respiratory Review 2024/25*. Available at: [asthmaandlung.org.uk/healthcare-professionals/ics-respiratory-review](https://www.asthmaandlung.org.uk/healthcare-professionals/ics-respiratory-review)

2.  UK Parliament. (2025). *Vaccination: Children – Written question 36734*. Available at: [questions-statements.parliament.uk/written-questions/detail/2025-03-10/36734](https://questions-statements.parliament.uk/written-questions/detail/2025-03-10/36734)

3.  Asthma + Lung UK. (2023). *Fighting to end the lung health lottery*. Available at: [asthmaandlung.org.uk/blog/campaigning/fighting-end-lung-health-lottery](https://www.asthmaandlung.org.uk/blog/campaigning/fighting-end-lung-health-lottery)

4.  World Health Organization. (2024). *Measles*. Available at: [who.int/news-room/fact-sheets/detail/measles](https://www.who.int/news-room/fact-sheets/detail/measles)

5.  Wellcome. (2026). *North and South: Regional Health Inequalities in England – Grants Awarded*. Available at: [wellcome.org/research-funding/funding-portfolio/funded-grants/north-and-south-regional-health-inequalities](https://wellcome.org/research-funding/funding-portfolio/funded-grants/north-and-south-regional-health-inequalities)

6.  Asthma + Lung UK. (2025). *ICS respiratory review – challenges*. Available at: [asthmaandlung.org.uk/healthcare-professionals/ics-respiratory-review/challenges](https://www.asthmaandlung.org.uk/healthcare-professionals/ics-respiratory-review/challenges)

------------------------------------------------------------------------

**Prepared by:** Marco Sorbona, PhD\
**Date:** March 2026\
**GitHub:** [MarcoSorbona/ukhsa-respiratory-and-outbreak-surveillance](https://github.com/MarcoSorbona/ukhsa-respiratory-and-outbreak-surveillance)\
**Live dashboard:** [marcoSorbona.github.io/ukhsa-respiratory-and-outbreak-surveillance/06_day6_dashboard.html](https://marcoSorbona.github.io/ukhsa-respiratory-and-outbreak-surveillance/06_day6_dashboard.html)

*Data sources: Day 1: Respiratory mortality (Indicator 40701); Day 2: C. difficile (ukhsadatR); Day 3: MMR coverage (Indicator 30309); Day 4: North‑South grouping of Day 1 data*
