---
title: "Respiratory Health Surveillance Dashboard"
author: "Marco Sorbona, PhD"
date: "March 2026"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme: cosmo
    source_code: embed
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(DT)
library(scales)
library(knitr)
library(geographr)
library(leaflet)

# Load data from previous days
respiratory_data <- readRDS("data/respiratory_data.rds") |> 
  filter(sex == "Persons") # Use overall rates only
north_south_data <- readRDS("data/north_south_data.rds")
mmr_data <- readRDS("data/mmr_data.rds")
cdiff_data <- readRDS("data/cdiff_data.rds")

# Load region boundaries
region_sf <- boundaries_region21

# Calculate key metrics
latest_year <- max(respiratory_data$year, na.rm = TRUE)

# Latest England rate
england_latest <- respiratory_data |> 
  filter(region == "England", year == latest_year) |> 
  pull(value) |> 
  first()

england_display <- paste0(round(england_latest, 1), "/100k")

# North-South difference
north_mean <- north_south_data |> 
  filter(region_group == "North") |> 
  pull(mortality_rate) |> 
  mean(na.rm = TRUE)

south_mean <- north_south_data |> 
  filter(region_group == "South") |> 
  pull(mortality_rate) |> 
  mean(na.rm = TRUE)

north_south_diff <- north_mean - south_mean
diff_display <- paste0("+", round(north_south_diff, 1), "/100k")

# Leicester MMR latest
leicester_mmr <- mmr_data |> 
  filter(geography == "Leicester", year == latest_year) |> 
  pull(coverage) |> 
  first()

mmr_display <- paste0(round(leicester_mmr, 1), " %")
mmr_color <- ifelse(leicester_mmr >= 95, "success", "warning")
```

# Overview {.tabset}

### Key Metrics

```{r}
# Create a simple data frame of metrics
metrics_table <- data.frame(
  Metric = c(
    "England respiratory mortality",
    "North vs South difference",
    "Leicester MMR coverage",
    "C. difficile alert rate"
  ),
  Value = c(
    paste0(round(england_latest, 1), " per 100k"),
    paste0("+", round(north_south_diff, 1), " per 100k"),
    paste0(round(leicester_mmr, 1), "%"),
    "80%"
  ),
  Year = c(
    latest_year,
    "2001–2024",
    latest_year,
    "2024"
  )
)

knitr::kable(metrics_table, caption = "Key Metrics")
```

### Data Summary

```{r}
# Create summary table of all datasets
data_summary <- data.frame(
  Dataset = c(
    "Respiratory Mortality",
    "North-South Grouped",
    "MMR Coverage",
    "C. difficile"
  ),
  `Time Period` = c(
    "2001-2024",
    "2001-2024",
    "2010-2024",
    "2023-2024"
  ),
  Geography = c(
    "England regions",
    "North vs South",
    "Leicester, East Midlands, England",
    "UKHSA regions"
  ),
  `Key Indicator` = c(
    "Rate per 100,000",
    "Rate per 100,000",
    "Percentage",
    "Weekly cases"
  ),
  check.names = FALSE
)

datatable(
  data_summary,
  options = list(
    pageLength = 5,
    dom = 't',
    searching = FALSE,
    paging = FALSE
  ),
  rownames = FALSE,
  class = "cell-border stripe hover"
)
```

------------------------------------------------------------------------

# Respiratory Mortality

### Time Series

```{r}
#| fig-height: 400


# Calculate regional summary for plot
regional_summary <- respiratory_data |> 
  mutate(
    region = str_remove(region, " region \\(statistical\\)")
    ) |> 
  group_by(region, year) |> 
  summarise(
    avg_rate = mean(value, na.rm = TRUE),
    .groups = "drop"
  )

p_ts <- ggplot(
  regional_summary,
  aes(x = year, y = avg_rate, color = region,
      text = paste("Region: ", region,
                   "<br>Year: ", year,
                    "<br>Rate: ", round(avg_rate, 1)))
) +
  geom_line(aes(group = region), linewidth = 0.8) +
  geom_point(size = 1) +
  scale_colour_viridis_d() +
  labs(
    title = "Respiratory mortality trends by region",
    x = "Year",
    y = "Rate per 100,000",
    color = ""
  ) +
    theme_minimal() +
    theme(legend.position = "bottom")

ggplotly(p_ts, tooltip = "text") |> 
  layout(
    legend = list(
      orientation = "h", 
      xanchor = "center", 
      x = 0.5, 
      y = -0.2, 
      font = list(size = 10)))
```

### Regional Map

```{r}
#| fig-height: 400


latest_year <- max(respiratory_data$year, na.rm = TRUE)
latest_regional <- respiratory_data |> 
  filter(year == latest_year) |> 
  mutate(
    region = str_remove(region, " region \\(statistical\\)")
  ) |> 
  arrange(desc(value))

p_rm <- ggplot(
  latest_regional,
  aes(x = reorder(region, value), y = value, fill = region)
) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(
    title = paste("Respiratory mortality by region (", latest_year, ")", sep = ""),
    x = "",
    y = "Rate per 100,000"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p_rm, tooltip = "text")
```

------------------------------------------------------------------------

# Map

### Respiratory Morality Interactive Map

```{r}
latest_year <- max(respiratory_data$year, na.rm = TRUE)

# Join mortality data with boundaries
map_data <- region_sf |>
  mutate(
    region_clean = str_remove(region21_name, " region \\(statistical\\)")
  ) |>
  left_join(
    respiratory_data |>
      filter(year == latest_year) |>
      select(region, value) |>
      mutate(
        region_clean = str_remove(region, " region \\(statistical\\)")
      ),
    by = "region_clean"
  )

# Create color palette
pal <- colorNumeric(
  palette = "viridis",
  domain = map_data$value,
  na.color = "#808080"
)

# Create interactive map
leaflet(map_data) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal(value),
    weight = 1,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 3,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0(
      region21_name, ": ",
      round(value, 1), " per 100,000"
    ),
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "10px",
      direction = "auto"
    )
  ) |>
  addLegend(
    pal = pal,
    values = ~value,
    opacity = 0.7,
    title = paste0(latest_year, " mortality rate<br>per 100,000"),
    position = "bottomright"
  ) |>
  setView(lng = -2.5, lat = 53.5, zoom = 6)
```

------------------------------------------------------------------------

# C. difficile Surveillance

### Outbreak Detection

```{r}

hospital_cases <- cdiff_data |> 
  filter(onset_type == "Hospital-onset, healthcare associated") |> 
  mutate(region = as.character(region))

hospital_summary <- hospital_cases |> 
  group_by(region, year, week) |> 
  summarise(
    cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  filter(!is.na(year))

p_od <- ggplot(
  hospital_summary,
  aes(x = week, y = cases, color = factor(year))
  ) +
  geom_line(aes(group = year), linewidth = 0.5, alpha = 0.7) +
  geom_point(size = 1) +
  facet_wrap(~region, scales = "free_y") +
  scale_color_manual(values = c("2023" = "#2c3e50", "2024" = "#e74c3c")) +
  labs(
    title = "Hospital-onset C. difficile cases by week",
    subtitle = "2023 vs 2024: each panel shows one region",
    x = "Week number",
    y = "Number of cases",
    color = "Year"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"))


ggplotly(p_od, tooltip = c("week", "year", "cases")) |> 
  layout(
    legend = list(
      orientation = "h",
      xanchor = "center",
      x = 0.5,
      y = -0.1,
      font = list(size = 10)
    )
  )
```

### Regional Heatmap

```{r}

hospital_heatmap <- hospital_cases |> 
  group_by(region, week) |> 
  summarise(
    total_cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  group_by(region) |> 
  mutate(
    region_mean = mean(total_cases, na.rm = TRUE),
    alert_strength = ifelse(total_cases > region_mean * 1.5, total_cases, 0)
  )

p_rhm <- 
  ggplot(
    hospital_heatmap,
    aes(x = week, y = region, fill = alert_strength,
        text = paste("Region: ", region,
                     "<br>Week: ", week,
                     "<br>Alert strength: ", alert_strength))
  ) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#e74c3c", name = "Alert\nstrength") +
  labs(
    title = "C. diffile outbreak signals by region",
    x = "Week",
    y = ""
  ) + 
  theme_minimal()

ggplotly(p_rhm, tooltip = "text")
```

------------------------------------------------------------------------

# MMR Vaccination

### Leicester Trends

```{r}

mmr_key <- mmr_data |> 
  filter(geography %in% c("Leicester", "East Midlands (average)", "England"))

p_lt <- 
  ggplot(
    mmr_key,
    aes(x = year, y = coverage, color = geography,
        text = paste("Geopraphy: ", geography,
                     "<br>Year: ", year,
                     "<br>Coverage: ", round(coverage, 1), "%"))
  ) +
  geom_hline(yintercept = 95, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_line(aes(group = geography), linewidth = 0.5) +
  geom_point(size = 1) +
  scale_color_manual(values = c(
    "Leicester" = "#3498db",
    "East Midlands (average)" = "#e67e22",
    "England" = "#2c3e50"
  )) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"), 
    limits = c(85,100)
  ) +
  labs(
    title = "MMR vaccination coverage by age 2",
    subtitle = "Red dashed line = WHO 95% target",
    x = "Year",
    y = "Coverage (%)",
    color = ""
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p_lt, tooltip = "text") |>
  layout(legend = list(
    orientation = "h",
    xanchor = "center",
    x = 0.5,
    y = -0.2
  ))
```

### Regional Benchmark

```{r}

latest_mmr <- mmr_data |> 
  filter(year == max(year)) |> 
  arrange(desc(coverage))

p_rb <- 
  ggplot(
    latest_mmr,
    aes(
      x = reorder(geography, coverage),
      y = coverage,
      fill = geography == "Leicester",
      text = paste("Geography: ", geography,
                   "<br>Coverage", round(coverage, 1), "%")
    )
  ) +
  geom_hline(yintercept = 95, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_col() +
  geom_text(aes(label = paste0(coverage, "%")), hjust = -0.1) +
  scale_fill_manual(values = c("TRUE" = "#3498db", "FALSE" = "grey70")) +
  coord_flip() +
  labs(
    title = paste("MMR coverage by geography (", max(latest_mmr$year), ")", sep = ""),
    subtitle = "Leicester highlighted in blue",
    x = "",
    y = "Coverage (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p_rb, tooltip = "text")
```

------------------------------------------------------------------------

# North-South Comparison

### Statistical Comparison

```{r}
p_sc <- 
  ggplot(
    north_south_data,
    aes(x = region_group, y = mortality_rate, fill = region_group,
        text = paste("Group: ", region_group,
                     "<br>Rate: ", round(mortality_rate, 1)))
  ) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c(
    "North" = "#e74c3c",
    "South" = "#3498db"
  )) +
  labs(
    title = "Distribution of respiratory mortality rates",
    subtitle = paste("North mean:", round(mean(north_south_data$mortality_rate[north_south_data$region_group == "North"], na.rm = TRUE), 1),
                     "| South mean:", round(mean(north_south_data$mortality_rate[north_south_data$region_group == "South"], na.rm = TRUE), 1)),
    x = "",
    y = "Rate per 100,000",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p_sc, tooltip = "text")
```

### Time Series with CI

```{r}

annual_ci <- north_south_data |> 
  group_by(region_group, year) |> 
  summarise(
    mean_rate = mean(mortality_rate, na.rm = TRUE),
    se = sd(mortality_rate, na.rm = TRUE) / sqrt(n()),
    ci_lower = mean_rate - 1.96 * se,
    ci_upper = mean_rate + 1.96 * se,
    .groups = "drop"
  )

p_ts_ci <-
  ggplot(
    annual_ci,
    aes(x = year, y = mean_rate, color = region_group, fill = region_group,
        text = paste("Group: ", region_group,
                     "<br>Year: ", year,
                     "<br>Rate: ", round(mean_rate, 1),
                     "<br>95% CI: [", round(ci_lower, 1),", ", round(ci_upper, 1), "]"))
  ) +
    geom_ribbon(
      aes(group = region_group, ymin = ci_lower, ymax = ci_upper),
      alpha = 0.1,
      color = NA
    ) +
    geom_line(aes(group = region_group), linewidth = 0.7) +
    geom_point(size = 1) +
    scale_color_manual(values = c("North" = "#e74c3c", "South" = "#3498db")) +
    scale_fill_manual(values = c("North" = "#e74c3c", "South" = "#3498db")) +
    labs(
      title = "North-South divide with 95% confidence bands",
      x = "Year",
      y = "Rate per 100,000",
      color = "",
      fill = ""
      ) +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  ggplotly(p_ts_ci, tooltip = "text") |> 
    layout(legend = list(orientation = "h", xanchor = "center", x = 0.5, y = -0.2))
```

------------------------------------------------------------------------

# Data Explorer {.tabset}

### Respiratory data

```{r}
datatable(
  respiratory_data |> 
    select(region, year, value, cases = count, population = denominator) |> 
    arrange(region, year) |> 
    mutate(
      region = str_remove(region, " region \\(statistical\\)"),
      across(where(is.numeric), ~round(., 1)),
      cases = format(cases, big.mark = ","),
      population = format(population, big.mark = ",")
      ),
  options = list(
    pageLength = 20,
    scrollX = TRUE,
    dom = 'Bfrtip'
  ),
  caption = "Respiratory mortlaity by region and year"
)
```

### MMR Data

```{r}
datatable(
  mmr_data |>
    select(geography, year, coverage_pct= coverage) |>
    arrange(geography, year) |>
    mutate(coverage_pct = round(coverage_pct, 1)),
  options = list(
    pageLength = 20,
    scrollX = TRUE,
    dom = 'Bfrtip'
  ),
  caption = "MMR coverage (%) by geography and year"
)
```

------------------------------------------------------------------------

# References

### Data Sources

| Source | Description | Access Date |
|------------------|---------------------------|---------------------------|
| **Fingertips** | Public health data for England | March 2026 |
| **UKHSA Data Dashboard** | Infectious disease surveillance | March 2026 |
| **NHS Digital** | Hospital Episode Statistics | March 2026 |
| **Office for National Statistics** | Population estimates | March 2026 |

### Packages Used

| Package         | Version | Purpose                    |
|-----------------|---------|----------------------------|
| `fingertipsR`   | 0.3.0   | Access to Fingertips data  |
| `ukhsadatR`     | 0.2.0   | UKHSA data API             |
| `tidyverse`     | 2.0.0   | Data wrangling             |
| `plotly`        | 4.10.0  | Interactive visualisations |
| `flexdashboard` | 0.6.0   | Dashboard layout           |
| `DT`            | 0.32.0  | Interactive tables         |
| `effectsize`    | 0.8.0   | Effect size calculations   |

### Reproducibility

All analyses can be reproduced by running the `00-setup.qmd` script, which downloads all data and saves it as RDS files. The dashboard loads these saved files, ensuring consistency across days.

**GitHub Repository:** [github.com/MarcoSorbona/ukhsa-respiratory-surveillance](https://github.com/MarcoSorbona/ukhsa-respiratory-surveillance)

### Software Environment

-   **R version:** 4.5.3
-   **RStudio version:** 2024.12.0+467
-   **Operating system:** Windows 11
